---
title: "cyclistic"
author: "ian.macdougall"
date: "10/29/2021"
output:
  pdf_document: default
editor_options:
  chunk_output_type: console
---

###load libraries
```{r}
library(data.table)
setDTthreads(4)
library(ggplot2)
library(ggridges)
```

###import data
```{r}
t.1 <- fread("~/R/capstone/cyclistic/cyclistic.data/trips.2019.q2.csv")
t.2 <- fread("~/R/capstone/cyclistic/cyclistic.data/trips.2019.q3.csv")
t.3 <- fread("~/R/capstone/cyclistic/cyclistic.data/trips.2019.q4.csv")
t.4 <- fread("~/R/capstone/cyclistic/cyclistic.data/trips.2020.q1.csv")

```

###convert to data.table
```{r}
t.1 <- as.data.table(t.1)
t.2 <- as.data.table(t.2)
t.3 <- as.data.table(t.3)
t.4 <- as.data.table(t.4)

is.data.table(t.1)
is.data.table(t.2)

```

###  clean data and ready to merge
  remove gender and birth year (mostly NAs)
  goal colnames = 
  c("trip.id", "start", "end", "seconds", "start.id", "start.name", "end.id", "end.name", "user.type")
  
###t.1  
```{r}
t.1[, `Member Gender` := NULL]
t.1[, `05 - Member Details Member Birthday Year` := NULL]
t.1[, '01 - Rental Details Bike ID' := NULL]

colnames(t.1) <- c("trip.id", "start", "end", "seconds", "start.id", "start.name", "end.id", "end.name", "user.type")

```

###t.2
```{r}
t.2[, 'gender' := NULL]
t.2[, 'birthyear' := NULL]
t.2[, 'bikeid' := NULL]

colnames(t.2) <- c("trip.id", "start", "end", "seconds", "start.id", "start.name", "end.id", "end.name", "user.type")

```

###t.3
```{r}
t.3[, 'gender' := NULL]
t.3[, 'birthyear' := NULL]
t.3[, 'bikeid' := NULL]

colnames(t.3) <- c("trip.id", "start", "end", "seconds", "start.id", "start.name", "end.id", "end.name", "user.type")

```

###t.4   (super messy)
```{r}
t.4[, 'start_lat' := NULL]
t.4[, 'start_lng' := NULL]
t.4[, 'end_lat' := NULL]
t.4[, 'end_lng' := NULL]
t.4[, 'rideable_type' := NULL]

#add column 'seconds
time <- c(t.4$ended_at - t.4$started_at)
t.4[, 'seconds' := time]
t.4 <- t.4[, seconds := as.character(seconds)]
rm(time)

#reorganize column order
setcolorder(t.4, c(1, 2, 3, 9, 5, 4, 7, 6, 8))

#rename columns
colnames(t.4) <- c("trip.id", "start", "end", "seconds", "start.id", "start.name", "end.id", "end.name", "user.type")

```

###check for NA's
```{r}
any(is.na(t.1))
any(is.na(t.2))
any(is.na(t.3))
any(is.na(t.4))  #true! find it!

sum(is.na(t.4))  # 1 NA somewhere
## row 414427

#delete NA row
t.4 <- na.omit(t.4)

```

###merge into a single DT
```{r}
l <- list(t.1, t.2, t.3, t.4)
trips <- rbindlist(l)
#rm(t.1, t.2, t.3, t.4)

```

###more cleaning
  seconds is character, convert to numeric (factor first?)
```{r}
trips <- trips[ , seconds := as.factor(seconds)]
trips <- trips[ , seconds := as.numeric(seconds)]
```


###user.type needs cleaning. want casual, member
```{r}
trips <- trips[user.type == "Customer", user.type := "casual"]
trips <- trips[user.type == "Subscriber", user.type := "member"]

# burn trip.id, start & end locations
trips <- trips[, ':='(trip.id = NULL, start.name = NULL, 
              start.id = NULL, end.id = NULL, end.name = NULL)]
```


###add columns for minutes, hours, weekday
```{r}
trips <- trips[ , minutes := round((seconds / 60))]

trips <- trips[ , hours := round((seconds / 3600))]

trips <- trips[, day := as.numeric(wday(trips$start))]
trips <- trips[, day := as.character(wday(trips$start))]
trips <- trips[day == 1, day := "sunday"]
trips <- trips[day == 2, day := "monday"]
trips <- trips[day == 3, day := "tuesday"]
trips <- trips[day == 4, day := "wednesday"]
trips <- trips[day == 5, day := "thursday"]
trips <- trips[day == 6, day := "friday"]
trips <- trips[day == 7, day := "saturday"]

setcolorder(trips, c("user.type", "start", "end", "seconds", "minutes", "hours", "day"))

```


###number of rides, per day of week, member/casual
```{r}
freq.day.mem <- trips[user.type == "member", list(count = .N), by = day]
freq.day.mem <- freq.day.mem[, user.type := "member"]

freq.day.cas <- trips[user.type == "casual", list(count = .N), by = day]
freq.day.cas <- freq.day.cas[, user.type := "casual"]

freq.day.user <- rbind(freq.day.cas, freq.day.mem)
freq.day.user$day.num <- rep(1:7, 2)

rm(freq.day.cas, freq.day.mem)

```


###average duration member/casual
```{r}
dmem <- trips[user.type == "member",
     list(mean = mean(minutes), 
     median = median(minutes)), 
     by = day]
dmem <- dmem[, user.type := "member"]
dmem$day.num <- rep(1:7)

dcas <- trips[user.type == "casual",
     list(mean = mean(minutes), 
     median = median(minutes)), 
     by = day]
dcas <- dcas[, user.type := "casual"]
dcas$day.num <- rep(1:7)


#duser <- rbind(dcas, dmem)
```


###plot median
```{r}
plot(x = dcas$day.num, y = dcas$median, t = "b", ylim = c(155, 265), col = "red",
     main = "Median Duration",
     ylab = "Duration (minutes)",
     xlab = "Day of Week",
     xaxt = "n")
axis(1, at = c(1:7), labels = c("M", "T", "W", "R", "F", "Sa", "Su"))
points(x = dmem$day.num, y = dmem$median, t = "b", col = "blue")
legend("left", legend = c("Member", "Casual"), col = c("blue", "red"), lwd = 1, bty = "n")
```

###plot mean
```{r}
plot(x = dcas$day.num, y = dcas$mean, t = "b", ylim = c(155, 265), col = "red",
     main = "Mean Duration",
     ylab = "Duration (minutes)",
     xlab = "Day of Week",
     xaxt = "n")
axis(1, at = c(1:7), labels = c("M", "T", "W", "R", "F", "Sa", "Su"))
points(x = dmem$day.num, y = dmem$mean, t = "b", col = "blue")
legend("bottomleft", legend = c("Member", "Casual"), col = c("blue", "red"), lwd = 1, bty = "n")

```


###plot rides per day
```{r}
fcas <- freq.day.user[user.type == "casual", .SD, by = day.num]
fmem <- freq.day.user[user.type == "member", .SD, by = day.num]

plot(x = fmem$day.num, y = fmem$count, 
     t = "b", 
     xlim = c(1, 7),
     main = "Rides per Day",
     xlab = "Day of Week", 
     ylab = "Number of Rides",
     ylim = c(90000,510000), 
     col = "blue",
     xaxt = "n")
axis(1, at = c(1:7), labels = c("M", "T", "W", "R", "F", "Sa", "Su"))
points(x = fcas$day.num, y = fcas$count, t = "b", xlim = c(1, 7), col = "red")
legend("left", legend = c("Member", "Casual"), col = c("blue", "red"), lwd = 1, bty = "n")
#dev.off()
```














